---
title: "Reproducible Research Assignment 1"
author: "Christian Lindenau"
date: "1/13/2022"
output: 
  html_document:
    css: mystyle.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

Loading packages

```{r load-packages, echo=TRUE, include=TRUE, message=FALSE, warning=FALSE}
library("dplyr")
library("tidyverse")
library("knitr")
library("ggplot2")
```

The acticity.csv dataset needs to be in a data subfolder of the working directory to read it

```{r, echo=TRUE}
data <- read.csv("./data/activity.csv")
```

Format the date column to a date format in R

```{r, echo= TRUE}
data$date <- as.Date(data$date, format = "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day

```{r, echo= TRUE}
aggday <- aggregate(data["steps"], by=data["date"], sum, na.rm = TRUE)
```

```{r, results='asis'}
writeLines("td, th { padding : 6px } th { background-color : brown ; color : white; border : 1px solid white; } td { color : brown ; border : 1px solid brown }", con = "mystyle.css")
knitr::kable(aggday, format = "html")
```

2. Make a histogram of the total number of steps taken each day

```{r, echo= TRUE}
phist <- aggday %>%
  ggplot( aes(x=steps)) +
  geom_histogram(bins = 30)
phist
```

3. Calculate and report the mean and median of the total number of steps taken per day and format as non-scientific number

```{r, echo= TRUE}
datamean <- format(mean(aggday$steps, na.rm = TRUE), scientific = FALSE)
datamedian <- format(median(aggday$steps, na.rm = TRUE), scientific = FALSE)
```

* Mean of the total number of steps taken per day: `r datamean`
* Median of the total number of steps taken per day: `r datamedian`

## What is the average daily activity pattern?

1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo= TRUE}
agginterval <-aggregate(data["steps"], by=data["interval"], mean, na.rm = TRUE)
ptimeseries <- agginterval %>%
  ggplot( aes(x=interval, y=steps)) +
  geom_line()
ptimeseries
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo= TRUE}
maxinterval <- slice_max(agginterval, steps)
```

The 5-minute interval `r maxinterval$interval` contains the maximum number of steps on average across all the days in the dataset. It contains `r maxinterval$steps` steps on average across all the days.

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset.

```{r, echo= TRUE}
countna <- sum(is.na(data$steps))
```

There are `r countna` missing values of steps in the dataset.

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

We will replace all missing values by the mean for that 5-minute interval

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r, echo= TRUE}
dataimputed <- data %>%
  group_by(interval) %>%
  mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = TRUE), steps))
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Aggregate the sum of the steps of the new imputed dataset by day

```{r, echo= TRUE}
aggimputedday <- aggregate(dataimputed["steps"], by=dataimputed["date"], sum, na.rm = TRUE)
```

Make a histogram of the total number of steps taken each day

```{r, echo= TRUE}
phistimputed <- aggimputedday %>%
  ggplot( aes(x=steps)) +
  geom_histogram(bins = 30)
phistimputed
```

Calculate and report the mean and median total number of steps taken per day

```{r, echo= TRUE}
dataimputedmean <- format(mean(aggimputedday$steps, na.rm = TRUE), scientific = FALSE)
dataimputedmedian <- format(median(aggimputedday$steps, na.rm = TRUE), scientific = FALSE)
```

* Mean of the total number of steps taken per day with missing data filled in: `r dataimputedmean`
* Median of the total number of steps taken per day with missing data filled in: `r dataimputedmedian`

Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

The values differ. Mean and Median have the same values because of the strategy for filling in the missing values.  

* Mean with missing data: `r datamean`, Mean without missing data: `r dataimputedmean`
* Median with missing data: `r datamedian`, Median without missing data: `r dataimputedmedian`

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

For aggregation per interval the dataset will be splitted into weekdays and weekend and later merged when aggregated per interval

```{r, echo= TRUE}
dataimputedweek <- cbind(dataimputed, weekday = !(weekdays(dataimputed$date) %in% c('Samstag' , 'Sonntag'))) %>%
  filter(weekday == TRUE)
dataimputedwknd <- cbind(dataimputed, weekday = !(weekdays(dataimputed$date) %in% c('Samstag' , 'Sonntag'))) %>%
  filter(weekday == FALSE)
intervalweek <- aggregate(dataimputedweek["steps"], by=dataimputedweek["interval"], mean)
intervalweek <- cbind(intervalweek, day = "weekday")
intervalwknd <- aggregate(dataimputedwknd["steps"], by=dataimputedwknd["interval"], mean)
intervalwknd <- cbind(intervalwknd, day = "weekend")
compareinterval <- merge(intervalweek, intervalwknd, all = TRUE)
```

2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data. 

```{r, echo= TRUE}
compareplot <- ggplot(data=compareinterval, aes(x=interval, y=steps)) + 
  geom_line() +
  facet_wrap(~day, dir = "v") +
  labs(title = "Weekdays vs. weekend", 'dir = "v"',
       y = "Number of steps")
compareplot
```

